\chapter*{Synthesis}
% Ajouter manuellement cette section à la table des matières
\addcontentsline{toc}{chapter}{Synthesis}

\subsection*{Introduction and Regulatory Context}
The Solvency 2 directive requires insurance undertakings to perform a rigorous assessment of their financial liabilities, necessitating the systematic use of stochastic Asset-Liability Management (ALM) models. These projections require considerable computing power, raising major operational challenges related to execution times. The central problem of this study consists in reconciling the reduction of the liability portfolio's dimension, through an aggregation process, with the preservation of the fidelity of regulatory risk indicators during sensitivity analyses. The study is based on a simulated reference portfolio of 1,000,000 policies, generated to faithfully reflect the technical characteristics of the French life insurance market.

\subsection*{Modeling Scope and Risk Indicators}
To measure the impact of aggregation on the risk profile, the analysis focuses primarily on the accuracy of two major regulatory indicators: the \textit{Best Estimate} (BE) and the Solvency Capital Requirement (SCR).

The evaluation of these metrics is carried out through a progressive complexification of the modeling chain. First, a liability-only model is used to isolate the natural runoff of flows and claims. Next, a deterministic ALM model is introduced to integrate simple financial revaluation, followed by the full stochastic ALM model. The latter is essential for understanding the non-linear interactions between financial returns and policyholder behavior, such as the exercise of surrender options, the de-sensitization of Minimum Guaranteed Rates (MGR), or the activation of profit sharing.

Valuation requires the integration of various economic environments. To this end, the work relies on the risk-free rate curves published by EIOPA, including the Volatility Adjustment (VA). For the measurement of regulatory sensitivities, three distinct economic situations are modeled using the outputs of an Economic Scenario Generator (ESG): the central stochastic trajectory, as well as two parallel shocks applied to the reference rate curve, amounting to $+100$ and $-100$ basis points, respectively.

\subsection*{Aggregation Principle and Definition of \textit{Model Points}}
Given the massive volume of individual contracts in the reference portfolio, line-by-line evaluation is technically prohibitive for production-level stochastic calculations. The standard optimization technique consists in grouping policies with similar risk profiles into reduced-size virtual portfolios.

Each grouping forms a \textit{Model Point} (MP), defined as a synthetic contract whose characteristics (such as age, policy tenure, or technical rate), weighted by mathematical provisions, aim to replicate the financial behavior of the underlying contracts. The primary technical challenge lies in the method of selecting and constituting these \textit{Model Points}. Excessive aggregation risks smoothing out specific guarantees, thereby altering the representation of the balance sheet's responsiveness to asymmetric market shocks and ultimately distorting the capital requirement measurement.

\subsection*{Stochastic Generation of the Reference Portfolio}
In the absence of access to unit-level data from a real portfolio for confidentiality reasons, the study relies on the creation of a synthetic portfolio of 1,000,000 contracts. To guarantee the representativeness of this liability portfolio, a stochastic policy generator was developed. The latter is based on the calibration of probability laws using public French market data (ACPR, INSEE, France Assureurs, general conditions of life insurance contracts).

The generation is structured around a generation-based approach, ensuring the temporal and financial consistency of the simulated technical characteristics. For each modeled random variable, a comparative statistical adjustment process was conducted, consisting in calibrating and comparing several theoretical probability distributions. The distributions finally selected are those demonstrating the best fit with empirical market observations:
\begin{itemize}
    \item \textbf{Policyholder age:} Modeled by a Beta distribution, specifically calibrated to reproduce the asymmetric demographic structure of life insurance holders;
    \item \textbf{Gender:} Generated via a Bernoulli distribution, parameterized according to the observed market distribution;
    \item \textbf{Contract seniority:} Simulated by a \textit{Gaussian Mixture Model} (GMM). This selection proved essential to faithfully reproduce various historical waves of subscription (underwriting shocks);
    \item \textbf{Mathematical Provisions (MP):} Outstanding amounts are generated by a Log-Normal distribution, capturing the asymmetry of wealth distribution and the existence of high-capital contracts;
    \item \textbf{Conditional technical characteristics:} Financial parameters, such as the Minimum Guaranteed Rate (MGR), the proportion of unit-linked (UL) accounts, and fee structures, are not drawn independently. They are assigned conditionally based on the year of subscription, reflecting the historical reality of regulatory technical rates and evolving commercial offers.
\end{itemize}

This simulation methodology allows for the restitution of the strong heterogeneity and structural correlations that a real insurance liability portfolio could have.

\subsection*{Typology of Aggregation Methods}
To compare different logic for dimension reduction, a typology of the studied models is established. The reference approach (\textit{Baseline}) corresponds to strict aggregation based on identical characteristics, i.e., an exclusive grouping of policies strictly sharing the same technical attributes (age, seniority, MGR, etc.), thus inducing no information loss.

This baseline is compared to the standard deterministic method of \textit{Banding}, which mechanically introduces intra-class discrepancies, as well as unsupervised learning algorithms (\textit{Machine Learning}) and a dynamic method: 
\begin{itemize}
    \item \textbf{The \textit{K-Means} algorithm:} Based on centroid partitioning, it aims to minimize the intra-class variance of \textit{Model Points} using the static characteristics of contracts;
    \item \textbf{The \textit{DBSCAN} and \textit{HDBSCAN} algorithms:} Based on local density estimation, they allow for the isolation of atypical profiles (noise management). Processing discrete variables with these spatial models requires the prior implementation of a \textit{Dithering} technique (adding uniform white noise);
    \item \textbf{The \textit{Cash-Flow Matching}:} Unlike the previous approaches applied to static characteristics at the closing date, this dynamic technique uses the \textit{K-Means} algorithm to group policies presenting similar future cash flow runoff trajectories.
\end{itemize}

The protocol for selecting optimal hyperparameters relies on an exhaustive \textit{Grid Search}. For each algorithm, the objective is to identify the configuration offering the Pareto optimum. The latter is defined as the best compromise between a strict operational compression constraint (targeting between 800 and 1,200 \textit{Model Points}) and the minimization of the relative error on the \textit{Best Estimate} evaluated using the liability-only model.

\subsection*{Optimal Parameterization of Aggregation Models}
The grid search protocol isolated the configurations maximizing projection fidelity while respecting the compression constraint. Table \ref{tab:synthese_calibration_finale_en} summarizes the hyperparameters selected for each algorithm.

\begin{table}[H]
    \centering
    \renewcommand{\arraystretch}{1.4}
    \begin{tabular}{|l|l|r|r|}
        \hline
        \textbf{Method} & \textbf{Optimal Parameterization} & \textbf{MP Nb} & \textbf{Liability-Only Error} \\
        \hline
        Banding & 5-year / 6-year brackets & 1,172 & $+0.0292\%$ \\
        K-Means & $k=1,039$ & 1,039 & $+0.0035\%$ \\
        DBSCAN & $\epsilon=0.0177$, MS$=5$ & 1,047 & $+0.0124\%$ \\
        HDBSCAN & $c=10$, $s=5$, \textit{leaf} (wt$=50$) & 967 & $+0.0398\%$ \\
        \textit{Cash-Flow Matching} & \textit{Direct on flows (1 run)} & 1,000 & $-0.0049\%$ \\
        \textit{A posteriori} calibration & Triplet (MGR, AFM, ESGMF) & 222 & $0.0000\%$ \\
        \hline
    \end{tabular}
    \caption{Summary of models selected for sensitivity analysis under Solvency 2 shocks.}
    \label{tab:synthese_calibration_finale_en}
\end{table}

The liability-only results demonstrate excellent compression capabilities for spatial and dynamic algorithms, maintaining a marginal relative error on pure flow runoff. The \textit{a posteriori} calibration method is also included for comparison, illustrating the naive approach of post-generation averaging.

\subsection*{Results in Stochastic ALM Projection (Central Scenario)}
The models were transposed into the stochastic ALM environment. Table \ref{tab:resultats_central_alm_en} presents the relative errors on the \textit{Best Estimate} (BE) and the global SCR.

\begin{table}[H]
    \centering
    \renewcommand{\arraystretch}{1.4}
    \begin{tabular}{|l|r|r|r|r|}
        \hline
        \textbf{Aggregation Method} & \textbf{Central BE} & \textbf{BE Error} & \textbf{Global SCR} & \textbf{SCR Error} \\
        \hline
        Baseline (Reference) & 209.93 Bn\euro{} & \textit{Ref.} & 26.91 Bn\euro{} & \textit{Ref.} \\
        Dynamic Banding & 209.02 Bn\euro{} & $-0.43\%$ & 27.47 Bn\euro{} & $+2.10\%$ \\
        HDBSCAN & 209.05 Bn\euro{} & $-0.42\%$ & 27.47 Bn\euro{} & $+2.10\%$ \\
        Robust K-Means & 208.95 Bn\euro{} & $-0.46\%$ & 27.49 Bn\euro{} & $+2.17\%$ \\
        \textit{Cash-Flow Matching} & 208.41 Bn\euro{} & $-0.72\%$ & 27.70 Bn\euro{} & $+2.94\%$ \\
        \textit{A posteriori} calibration & 198.31 Bn\euro{} & $-5.53\%$ & 19.27 Bn\euro{} & $-28.38\%$ \\
        \hline
    \end{tabular}
    \caption{Global evaluation of liabilities and SCR in stochastic ALM projection.}
    \label{tab:resultats_central_alm_en}
\end{table}

The error decomposition across risk modules (Table \ref{tab:resultats_modules_en}) allows for the identification of the origin of aggregation biases.

\begin{table}[H]
    \centering
    \renewcommand{\arraystretch}{1.4}
    \begin{tabular}{|l|r|r|r|}
        \hline
        \textbf{Aggregation Method} & \textbf{Market SCR Error} & \textbf{Life SCR Error} & \textbf{Global SCR Error} \\
        \hline
        Dynamic Banding & $+1.89\%$ & $+3.55\%$ & $+2.10\%$ \\
        HDBSCAN & $+1.87\%$ & $+3.71\%$ & $+2.10\%$ \\
        Robust K-Means & $+1.92\%$ & $+3.87\%$ & $+2.17\%$ \\
        \textit{Cash-Flow Matching} & $+2.42\%$ & $+6.50\%$ & $+2.94\%$ \\
        \hline
    \end{tabular}
    \caption{Decomposition of the error on the main risk modules.}
    \label{tab:resultats_modules_en}
\end{table}

Spatial approaches (\textit{HDBSCAN}, \textit{K-Means}, \textit{Banding}) faithfully replicate the balance sheet's responsiveness to market shocks, maintaining an error rate below $2\,\%$. By isolating outlier profiles, particularly high Minimum Guaranteed Rates, they prevent the dilution of guarantees within clusters. Conversely, \textit{Cash-Flow Matching} exhibits a more significant deviation in the Life SCR ($+6{,}50\,\%$). Based exclusively on flow morphology, this approach overlooks tax seniority data, leading to a mechanical overvaluation of the capital requirement for surrender risk. Finally, \textit{a posteriori} calibration proves unsuitable for stochastic projections: the static nature of behavioral laws results in a global SCR underestimation of more than $28\,\%$.

\subsection*{Sensitivity Analysis Facing Asymmetric Environments}
The evaluation was subjected to interest rate curve shocks (EIOPA $\pm 100$ bps) to test the non-linear distortion of surrender risk (Table \ref{tab:sensibilites_scr_synthese_en}).

\begin{table}[H]
    \centering
    \renewcommand{\arraystretch}{1.4}
    \begin{tabular}{|l|r|r|r|}
        \hline
        \multirow{2}{*}{\textbf{Aggregation Method}} & \textbf{Down (-100 bps)} & \multicolumn{2}{c|}{\textbf{Up (+100 bps)}} \\
        \cline{2-4}
        & \textbf{Global SCR Error} & \textbf{Global SCR Error} & \textbf{incl. Surrender SCR Error} \\
        \hline
        Dynamic Banding & $+1.17\%$ & $+1.23\%$ & $+4.55\%$ \\
        HDBSCAN & $+1.16\%$ & $+1.24\%$ & $+4.68\%$ \\
        Robust K-Means & $+1.21\%$ & $+1.25\%$ & $+4.72\%$ \\
        \textit{Cash-Flow Matching} & $+1.39\%$ & $+1.27\%$ & $+6.23\%$ \\
        \hline
    \end{tabular}
    \caption{Relative errors on the Global SCR and Surrender SCR according to curve distortion.}
    \label{tab:sensibilites_scr_synthese_en}
\end{table}

Surrender dynamics are highly asymmetric. In a low-rate environment, asset yields struggle to cover MGRs; an early exit generates little loss of profit, limiting the SCR. During a sudden rate hike, yields far exceed guarantees. A policyholder's exit then deprives the company of significant future margins. It is the realization of this opportunity cost that causes the Surrender SCR to jump.

Spatial methods, by preserving the granularity of MGRs and taxation, assess this risk accurately (global error stabilized around $+1.25\%$ in the "Up" scenario). \textit{Cash-Flow Matching} shows a higher deviation on the Surrender SCR ($+6.23\%$), confirming the limits of approaches purely based on central flows for modeling extreme asymmetric behaviors.

\subsection*{Operational Performance and Conclusion}
The reduction in dimensionality offers a decisive performance gain, quantified in Table \ref{tab:runtimes_synthese_en}.

\begin{table}[H]
    \centering
    \renewcommand{\arraystretch}{1.4}
    \begin{tabular}{|l|r|r|r|}
        \hline
        \textbf{Method} & \textbf{Volume (MP)} & \textbf{Execution Time} & \textbf{Reduction Factor} \\
        \hline
        Baseline (Reference) & 26,272 & $\approx$ 8 h 00 (28,783 s) & \textit{Ref.} \\
        HDBSCAN & 967 & $\approx$ 10 min 00 (600 s) & $\div 48$ \\
        \textit{Cash-Flow Matching} & 1,000 & $\approx$ 10 min 27 (627 s) & $\div 46$ \\
        Robust K-Means & 1,039 & $\approx$ 11 min 11 (671 s) & $\div 43$ \\
        Dynamic Banding & 1,172 & $\approx$ 11 min 51 (711 s) & $\div 40$ \\
        \hline
    \end{tabular}
    \caption{Comparison of calculation times in full stochastic ALM projection.}
    \label{tab:runtimes_synthese_en}
\end{table}

Compression divides processing times by a factor of approximately 48, bringing an 8-hour projection down to about ten minutes. This gain removes a technological bottleneck, allowing for the multiplication of sensitivity calculations on standard infrastructures.

In conclusion, this study validates the use of unsupervised learning algorithms for aggregation in the framework of ALM modeling, while providing critical insight into market practices:

\begin{itemize}
    \item \textbf{The robustness of the empirical approach:} The standard \textit{Banding} method demonstrates excellent resilience facing asymmetric shocks. It confirms that static variables (MGR, seniority) are the optimal grouping axes. However, its rigid mesh limits its maximum compression capabilities.
    \item \textbf{The industrial optimum through \textit{Machine Learning}:} To overcome this compression threshold, unsupervised algorithms are essential. While \textit{HDBSCAN} offers remarkable theoretical precision through its density-based approach, its mechanics make it difficult to target a specific number of clusters. The \textit{K-Means} algorithm thus represents the best operational compromise: it provides direct control over the final count of \textit{Model Points} while accurately replicating the balance sheet's responsiveness to economic variations.
    \item \textbf{The limit of dynamic approaches:} The \textit{Cash-Flow Matching} method proves effective for central projections, but its smoothing of seniority penalizes it during the assessment of regulatory surrender shocks.
\end{itemize}

Looking ahead, the integration of individual taxation as an explicit dimension within dynamic models could correct the biases observed during mass surrenders. Furthermore, the intrinsic nature of the \textit{HDBSCAN} algorithm offers a particularly relevant research path: rather than constraining the final volume, this method could allow for the organic determination of the optimal number of \textit{Model Points} required to structure a portfolio. The deep integration of \textit{Machine Learning} algorithms thus opens a new dimension for risk management in life insurance, moving definitively away from arbitrary volumetric standards.